preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-31 Manuscript/manuscript.Rmd
editted results

@@

We developed a machine learning framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview]. We then developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples;!!!fixed]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of TADs, TAD boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2 [??? I can see numbers for K562 (?) only? There's a discrepancy with legend.!!!fixed, need to added spectralTAD]). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 [??? 77 in the table?;!!!fixed] cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

@@

In summary, we considered combinations of data from two cell line *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model.

@@

<!-- Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for Peakachu defined boundaries (Supplementary Tables 3-4). -->

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple random under-sampling can significantly improve TAD boundary region prediction.

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions. These results were consistent for the K562 cell line (Supplementary Figure 2). Random forest models built on Peakachu loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Five kilobase resolution genomic bins was found to result in optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and loop boundary prediction.

<!--
Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.
-->

@@

We hypothesized that the type of genomic annotations may also affect the predictive performance. Using the established optimal settings (RUS, Distance), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries, for both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

We sought to further optimize our TAD boundary region prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

<!--
To summarize, the optimally built cell line-specific random forest model under our computational framework was found to be one built using 10 kb resolution data, leveraging distance-type predictors for the CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites, while implementing random under-sampling to balance TAD and non-TAD boundary defined genomic bins.
-->

>>>>

## Evaluating performance across cell lines

When evaluating performance across cell lines, a random forest model was built using ground truth boundaries and genomic annotations from one cell line such that the data matrix was constructed using genomic bins from $n-1$ chromosomes. The test data used to evaluate performance was constructed from the binned data of the $i^{th}$ holdout chromosome with genomic annotations from the other cell line. Thus, the models were trained on cell line specific data and tested on the other cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivies and specificities when tested on the cell line specific holdout chromosome test data, and reported the area under the curve (AUC).

@@

We developed a machine learning framework aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an algorithm which leverages an optimally built RF to precisely predict the genomic coordinates of both TAD and chromatin loop boundaries at base pair resolution.

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loops, we opted to use boundaries obtained by the PEAKACHU prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

Boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell-line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 3). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These include signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods).

@@

We considered combinations of data from two cell line *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model.

@@

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types when predicting TAD boundary regions.

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions with results consistent for K562 (Supplementary Figure 2). Random forest models built on PEAKACHU-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Furthermore, 5 kb resolution genomic bins led to optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

@@

We hypothesized that the type of genomic annotations may also affect the predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562.

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome that was left out of training and reserved for testing. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, under our machine learning framework we were able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

## Training in one cell line accurately predicts boundary regions in other cell lines

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using ARROWHEAD ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on PEAKACHU resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D).
